---
title: "Modelling Missing Data"
author: "David Luke Thiessen"
date: "September 19, 2018"
output: beamer_presentation
bibliography: bibliography.bibtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
    - Four spaces makes double indent
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

## Another slide I made

A sample Latex Equation

\begin{equation}
X = \sum_{i = 1}^{5} y_i
\end{equation}

## Why does this look different?

More sample text [see @Nobody06].

The name of the reference is case sensitive. 

## References

