---
title: "Models for Non-Monotone Missing at Random Data"
author: "David Luke Thiessen"
date: "October 15, 2018"
output: 
  beamer_presentation:
   includes:
      in_header: headerpagenumbers.tex
bibliography: bibliography.bibtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Outline
- Introduction 
- Background on Missing Data Theory
- Approaches for Analyzing Missing Data
- Inverse Probability Weighting
- Methods for Modelling Missing Data Probabilities


## Introduction
In the basic statistical structure, there is a "population" consisting of a collection of individuals. We are interested in studying some aspect of the population (commonly a single characteristic or a relation between several characteristics), but for reasons of cost/time/convenience, are unable to take an observation from every individual.

Our approach is to randomly select a portion of the population, record observations from these individuals, then make generalizations to the rest of the population.


## Introduction: Example
A healthcare worker at a retirement home with 100 residents wants to know if the residents have high blood pressure. If they do, the retirement home should consider changing the menu to help improve their health. She randomly selects 30 residents to record their blood pressure.

The estimate of average blood pressure is:

\begin{equation}
\bar{x} = \frac{1}{n}\sum_{i=1}^{30} x_i
\end{equation}

and (under typical assumptions) the confidence interval would be:

\begin{equation}
(\bar{x} - z_{\alpha/2}\frac{s}{\sqrt{n}}, \, \bar{x} + z_{\alpha/2}\frac{s}{\sqrt{n}})
\end{equation}


## Introduction: The Problem
Some of the retirees refuse, and won't let the healthcare worker measure their blood pressure.

How should the worker calculate the estimate and confidence interval?

Remark: Missingness is a nuisance, not the objective of scientific study.

Question: Can auxillary information (like the age of the residents) be used to recover (or estimate) the missing information?


## Introduction: A Naive Approach
Suppose that the healthcare worker decides to just ignore the residents who refuse to have their blood pressure recorded. She considers them as if they were not selected to be in the sample and estimates the average as:

\begin{equation}
\bar{x} = \frac{1}{|J|}\sum_{i \in J} x_i
\end{equation}

Where J is the set of residents in the sample who let their blood pressure be recorded.

Will this approach give an accurate estimate?

Answer: It depends, but in general this estimate is biased.


## Background: Notation
As in @SunBaoluo2017OIPW, we let $X = (X_1, X_2, ..., X_K)$ be a random vector representing the complete (fully observed) data.

Let $R$ be an integer variable denoting which of the $X_i$ variables have been observed, with $1 \leq R \leq 2^K$. Let $R = 1$ indicate that all variables observed and $R = r > 1$ denote that only the variables $X_{(r)} \subset X$ are observed (which could be the empty set).

Each individual is an independent and identically distributed observation from the random variable $(R_i, X_{(R)i})$. 


## Background: Missing Data Patterns
If the variables $X_i$ can be arranged such that $X_{(r+1)} \subset X_{(r)}$, then the data are said to have a monotone missing pattern. This is common in long-term studies where participants drop out part-way through.

```{r}
X = c(6,4,4,3,2)
barplot(X,axes=FALSE,main="Monotone Missing",names.arg=c("X1","X2","X3","X4","X5"),density=c(rep(20,5)))
```


## Background: Missing Data Patterns
If the variables are not monotone missing, then they are non-monontone missing.

```{r}
X = c(6,4,4,3,2)
barplot(X,axes=FALSE,main="Non-monotone Missing",names.arg=c("X1","X2","X3","X4","X5"),density=c(rep(20,5)))
rect(xleft = 2.6, ybottom = 5, xright = 3.6, ytop = 6, density = 20,lwd=0.6)
rect(xleft = 5, ybottom = 3, xright = 6, ytop = 5, density = 20,lwd=0.6)
```


## Background: Missing Data Mechanisms
Missing Completely at Random (MCAR)

- Missingness is independent of the data.
- Example: The device the healthcare worker is using to record blood pressure is faulty, and the device randomly fails when trying to use it on some residents.

\begin{equation}
\pi_r(x) = P(R = r|x,\gamma) = P(R=r|\gamma)
\end{equation}


## Background: Missing Data Mechanisms
Missing at Random (MAR)

- Missingness can depend on data values that are observed, but not those that are unobserved.
- Example: The residents are more likely to refuse the healthcare worker the older they are (and the worker has a list of the ages of residents).

\begin{equation}
\pi_r(x) = P(R=r|x,\gamma) = P(R=r|x^{\text{obs}},\gamma)
\end{equation}


## Background: Missing Data Mechanisms
Missing Not at Random (MNAR)

- Neither of the two above conditions hold.
- Example: Residents with high blood pressure are more likely to refuse to have their blood pressure recorded.


## Background: Examples of Missing Data Mechanisms
Suppose that the data in question is $(X_1,X_2) \sim \text{N}(\mu,\Sigma)$, where $\mu = (1,1)$, $\Sigma = \bigl( \begin{smallmatrix} 1 & 0.5 \\ 0.5 & 1\end{smallmatrix}\bigr)$ and the research objective is to study the variable $X_2$.

Let $X_1$ be always observed, and $X_2$ partially missing with the mechanism given by:

- MCAR: $P(R=2|x,\gamma) = P(R=2) = 0.4$

- MAR: $P(R=2|x,\gamma) = P(R=2|x_1) = 1/(1+e^{x_1})$

- MNAR: $P(R=2|x,\gamma) = P(R=2|x_2) = 1/(1+e^{x_2})$


## Background: Missing Data Mechanisms
Remark: For technical reasons, we also need to make the assumption that:

\begin{equation}
\pi_1(x) = P(R = 1|x,\gamma) > \sigma > 0
\end{equation}

for some constant $\sigma$ and all possible $x$.


## Approaches for Analyzing with Missing Data
TODO: 

- Complete Case Analysis (CC)
    - Throw out the individuals with missing data, and do an analysis as if you had only measured n' individuals instead of the originally planned n.
- Inverse Probability Weighting (IPW)
    - Weight the different individuals differently in the analysis, making ones with rare observations carry more weight
- Single Imputation (SI)
    - Replace the missing values with other appropriately selected values.
    - Sometimes the average is used. In many situations this is wrong.
- Multiple Imputation (MI)
    - Replace missing values with randomly (but appropriately) selected values, but repeat this many times and then combine the conclusions together in a principled way.
    
- Likelihood Methods?


## Demonstration
TODO: Everything after here, use the MVN example.


## Show that CC is unbiased for MCAR


## Show that CC is biased for MAR and MNAR


## Demostrate how IPW can correct for MAR


## Show how IPW Improves MNAR


## Another example with MNAR where IPW cannot correct for bias.


## Single Imputation - Doesn't correct the bias, gives a false sense of security by reducing the variance


## Demonstrate how Multiple Impuation works?


## Inverse Probability Weighting
- Efficiency
- Easier to specify than full likelihood
- Less compuation demanding?


## Modelling Missing Data Probability
- Logistic
- Multinomial with only a few missing patterns and lots of data
- UMLE and CBE from Paper



## References

